### 2024-04-19 docs/SOSP 2024/intro.tex
Edits to intro and motivation

%Despite these recent advancements, one dominant family of models, namely decision forest models have not seen the same level of acceleration.
Decision forest models remain the mainstay for machine learning over tabular data~\cite{DLNotAllYouNeed,TreebasedOutperformDL}. 
Their robustness, interpretability, and ability to handle missing data make them a popular choice for a wide 
range of applications\cite{DecisionTreesOverview, Med1, Med2, Facebook, LHCModel, Finance}. 
% \TODO{a few more lines, classification, regression, etc.}
% We observe that while scalable performance on neural network models has received significant attention, decision tree inference has not seen the same level of acceleration.
A recent survey~\cite{KaggleSurvey} found that about three quarters of data scientists use decision tree
based models. An analysis of ML workloads at a large scale web company found that these models are most widely used~\cite{LookingGlass}. 
Recent work has noted that the cost of inference is the most critical factor in the overall cost of 
deploying a machine learning model~\cite{Hummingbird, SageMaker}.
This is because, in production settings, each model is trained once and often used for inference millions of times. 
%% RG: Minor reword
%% However, 
Further, inference is run on a variety of hardware platforms, ranging from low to high-end CPUs and GPUs. 
This paper is motivated by the need to accelerate decision tree inference to achieve portable performance on 
%% RG: reword -- specifically say CPUs and GPUs.
%% a variety of hardware.
commodity platforms with CPUs and GPUs. 
% In particular, we focus on a range of commodity CPUs and GPUs, a class of hardware that has seen widespread 
% adoption across client/edge devices used for inference.

@@

This paper presents \Treebeard{}, a novel schedule guided compilation infrastructure
for decision tree inference on multiple target hardware. \Treebeard{} is able to 
generate high-performance code for decision tree inference by exploring a large optimization space. 
%% RG - rewording
%% The code generation is guided by a \emph{schedule} written in a custom scheduling language that can represent a wide range of implementation strategies. 
This is achieved by a compilation framework consisting of a 
custom scheduling language that can represent a wide range of 
implementation strategies and techniques to efficiently explore
the optimization space. We demonstrate that the language is 
sufficient to express the various optimizations proposed by prior work
and that our schedule exploration heuristic can quickly find a near 
optimal schedule for the model being compiled. 
%% Further we generalizes them and incorporates several new optimizations. 
\TODO{RG: it would be good to expand on this and also talk about the retargetable component. 
We could also say that the schedule framework and the retargetable compiler work in an 
interwined manner, each benefitting from the other.}  
The second component of \Treebeard{} is a \emph{retargetable} multi-level compiler that can generate efficient code 
for any specified schedule for both CPUs and GPUs. For this purpose, we re-architect \TreebeardOLD{} to
support schedule-guided code generation, and incorporate several new optimizations. These two components 
of the proposed \Treebeard{} compiler infrastructure are interwined, each benefitting from the other.
%Last, to address the challenge of selecting a near-optimal schedule from the vast optimization space, 
%%
%we design and implement a heuristic that is able to quickly find high-performance schedules for the model being compiled.
\TODO{Performance evaluation summary}
%% RG: This may not be needed -- can be removed to save space!
%% \TODO{Oganization}
\subsection{Contributions}
\begin{itemize}
  \item We present the design of a multi-target compiler infrastructure for decision tree 
   inferencing  and implement several optimizations within this framework.
   We are also the first to implement an optimizing compiler for decision tree 
   inference on GPUs. \TODO{AP: Given that there is Hummingbird, can we really say this?}
  \item We identify that an extensive optimization space exists for the problem of 
  decision tree inference. We design a scheduling language that 
  allows us to effectively represent this solution space abstractly. This 
  scheduling language is expressive enough to represent a wide range of  
  implementation strategies,  %%.  proposed by prior work.
%% RG: added CPUs and GPUs. 
  such as different data layouts, caching strategies, parallel reduction schemes, 
  that work across CPUs and GPUs. 
  \item To the best of our knowledge, we perform the first extensive characterization
  of the optimization space for decision tree inference 
  on GPUs. Using some of the characteristics we identify, we design and implement 
  a heuristic that is able to quickly find high-performance 
  schedules for the model being compiled. 
  \item We design and implement a general framework for expressing and optimizing
  reductions within MLIR. To the best of our knowledge, this is the first
  such framework.
  \item We evaluate our implementation by comparing it against RAPIDs and Tahoe, 
  the state-of-the-art decision tree inference frameworks for GPU and 
  report significant speedups. We also show that our compiler can effectively 
  target different GPUs, including both NVIDIA and AMD GPUs.
\end{itemize}

>>>>

%Despite these recent advancements, one dominant family of models, namely decision forest models have not seen the same level of acceleration.
Decision forest models remain the mainstay for machine learning over tabular data~\cite{DLNotAllYouNeed,TreebasedOutperformDL}. 
Their robustness, interpretability, and ability to handle missing data make them a popular choice for a wide 
range of applications\cite{DecisionTreesOverview, Med1, Med2, Facebook, LHCModel, Finance}. 
% \TODO{a few more lines, classification, regression, etc.}
% We observe that while scalable performance on neural network models has received significant attention, decision tree inference has not seen the same level of acceleration.
The Kaggle AI report for 2023 \cite{ABCD} highlights the continuing \emph{dominance of 
gradient boosted trees as the algorithm of choice} for tabular data. It also estimates that \emph{between
50\% and 90\% of practicing data scientists use tabular data as their primary type of data in their professional setting.}
% An analysis of ML workloads at a large scale web company found that these models are widely used~\cite{LookingGlass}.
This has also been observed in other surveys~\cite{KaggleSurvey,LookingGlass}.
Recent work has noted that the cost of inference is the most critical factor in the overall cost of 
deploying a machine learning model~\cite{Hummingbird, SageMaker}.
This is because, in production settings, each model is trained once and often used for inference millions of times. 
%% RG: Minor reword
%% However, 
Further, inference is run on a variety of hardware platforms, ranging from low to high-end CPUs and GPUs. 
This paper is motivated by the need to accelerate decision tree inference to achieve portable performance on 
%% RG: reword -- specifically say CPUs and GPUs.
%% a variety of hardware.
commodity platforms with CPUs and GPUs. 
% In particular, we focus on a range of commodity CPUs and GPUs, a class of hardware that has seen widespread 
% adoption across client/edge devices used for inference.

@@

This paper presents \Treebeard{}, a novel schedule guided compilation infrastructure
for decision tree inference on multiple hardware targets. \Treebeard{} is able to 
generate high-performance code for decision tree inference by exploring a large optimization space. 
%% RG - rewording
%% The code generation is guided by a \emph{schedule} written in a custom scheduling language that can represent a wide range of implementation strategies. 
This is achieved by a compilation framework consisting of a 
custom scheduling language that can represent a wide range of 
implementation strategies and techniques to efficiently explore
the optimization space. We demonstrate that the language is 
is capable of expressing all optimizations proposed by prior work and more.
Further, our schedule exploration heuristic can quickly find a near 
optimal schedule for the model being compiled. 
%% Further we generalizes them and incorporates several new optimizations. 
% \TODO{RG: it would be good to expand on this and also talk about the retargetable component. 
% We could also say that the schedule framework and the retargetable compiler work in an 
% interwined manner, each benefitting from the other.}  
The second component of \Treebeard{} is a \emph{retargetable} multi-level compiler that can generate efficient code 
for any specified schedule for both CPUs and GPUs. For this purpose, we re-architect \TreebeardOLD{} to
support schedule-guided code generation, and incorporate several new optimizations that are critical to target GPUs. 
These two components of the proposed \Treebeard{} compiler infrastructure are intertwined, each benefitting from the other.
%Last, to address the challenge of selecting a near-optimal schedule from the vast optimization space, 
%%
%we design and implement a heuristic that is able to quickly find high-performance schedules for the model being compiled.

We evaluated \Treebeard{} on a large number of synthetic and real-world 
benchmark models and compared against state-of-the-art systems RAPIDS~\cite{RAPIDS}, Tahoe~\cite{Tahoe}
, XGBoost~\cite{XGBoost} and \TreebeardOLD{}~\cite{Treebeard}. While these systems can only run on 
specific hardware, \Treebeard{} outperforms all these systems while providing portable performance across 
a range of target hardware. The geomean speedup (across all benchmarks) of \Treebeard{} over
Tahoe and RAPIDS is in the range $2$--$5\times$ over several batch sizes. The geomean speedup over
XGBoost is more than $10\times$. \Treebeard{} also enables better scalability on CPUs by parallelizing 
across multiple dimensions. We also demonstrate that \Treebeard{}'s scheduling heuristic is able to 
find near-optimal schedules in less than 30 seconds on average.

%% RG: This may not be needed -- can be removed to save space!
%% \TODO{Oganization}
\begin{itemize}
  \item We identify that there are many different ways to implement  
  decision tree inference and that the best implementation depends both on the model
  and the target. We design a scheduling language that 
  allows us to represent this solution space. 
%   This 
%   scheduling language is expressive enough to represent a wide range of  
%   implementation strategies,  %%.  proposed by prior work.
% %% RG: added CPUs and GPUs. 
%   such as different data layouts, caching strategies, parallel reduction schemes, 
%   that work across CPUs and GPUs. 
  \item We propose a retargetable compiler that can generate code as specified by 
  the schedule. The compiler also includes new abstractions for caching,
  reduction and representing models in memory. These are critical to generate 
  efficient GPU code.
  \item The scheduling language can represent an unbounded number of schedules. 
  We design and implement techniques to search over this space and find high-performance 
  schedules in tens of seconds. 
  \item We exhaustively evaluate \Treebeard{} and report that it achieves 
  portable performance across a range of hardware targets and consistently outperforms
  state-of-the-art decision tree inference frameworks for both CPU and GPU.
\end{itemize}

### 2024-04-19 docs/SOSP 2024/schedule.tex
Edits to intro and motivation

In the \emph{shared data} strategy~\cite{Tahoe}, a thread block walks all the trees 
for a given row in parallel. 
% If threads walk multiple trees, each thread accumulates partial results. 
Then, a thread block wide reduction is performed to compute 
the prediction. The schedule for this strategy is:
\begin{lstlisting}[style=c++]
reorder(batch, tree)
gpuDimension(batch, grid.x)
gpuDimension(tree, block.x)
reduce(batch)
\end{lstlisting}
%   \item \textbf{Shared Forest}: In this strategy, the whole model is loaded into 
%   shared memory and subsequently, a single thread walks all trees for a particular
%   row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
%   tile(tree, t0, t1, N_TREES)
%   reorder(b0, b1, t0, t1)
%   cache(t0)
%   gpuDimension(b0, grid.x)
%   gpuDimension(b1, block.x)
% \end{lstlisting}
%   Here, we create a placeholder single iteration loop \op{t0} so that we can 
%   specify that all trees are to be cached.
%   \item \textbf{Shared Partial Forest}: In case the model is too large to fit into
%   shared memory, the model is split into chunks and each chunk is loaded into shared
%   memory. Again, as in the previous strategy, one thread walks all trees assigned to
%   a thread block for a row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)

>>>>

In the \emph{shared data} strategy~\cite{Tahoe}, a thread block walks all the trees 
for a given row in parallel. Input rows are cached in shared memory.
% If threads walk multiple trees, each thread accumulates partial results. 
% Then, a thread block wide reduction is performed to compute 
% the prediction.
The schedule for this strategy is:
\begin{lstlisting}[style=c++]
reorder(batch, tree)
gpuDimension(batch, grid.x)
gpuDimension(tree, block.x)
cache(batch)
\end{lstlisting}
%   \item \textbf{Shared Forest}: In this strategy, the whole model is loaded into 
%   shared memory and subsequently, a single thread walks all trees for a particular
%   row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
%   tile(tree, t0, t1, N_TREES)
%   reorder(b0, b1, t0, t1)
%   cache(t0)
%   gpuDimension(b0, grid.x)
%   gpuDimension(b1, block.x)
% \end{lstlisting}
%   Here, we create a placeholder single iteration loop \op{t0} so that we can 
%   specify that all trees are to be cached.
%   \item \textbf{Shared Partial Forest}: In case the model is too large to fit into
%   shared memory, the model is split into chunks and each chunk is loaded into shared
%   memory. Again, as in the previous strategy, one thread walks all trees assigned to
%   a thread block for a row. The schedule for this strategy is as follows.
% \begin{lstlisting}[style=c++]
%   tile(batch, b0, b1, ROWS_PER_TB)
